---
title: "HPV-induced host epigenetic reprogramming is lost upon progression to high-grade cervical intraepithelial neoplasia"
author: "Main Figures"
date: ""
output:
  pdf_document:
    toc: false
    latex_engine: xelatex
header-includes:
  \usepackage{fontspec}
  \usepackage[utf8]{inputenc}
  \usepackage{pdfpages}
  \usepackage{graphicx}
  \setmainfont{Arial}
---

```{r, echo=FALSE, message = FALSE, warning = FALSE}
library(knitr)
library(ggplot2)
library(patchwork)
library(dplyr)
library(tidyverse)
library(ggbeeswarm)
library(gghalves)
library(pROC)
library(viridis)
library(ComplexHeatmap)
library(ggplotify)
library(ggpubr)
knitr::opts_chunk$set(echo=FALSE, message = FALSE, warning = FALSE)

theme_set(theme_minimal())

# load helper functions
source("../1-source-codes/plot_roc_dichot.R")
source("../1-source-codes/GOplot_methyl.R")

```


Chiara Herzog\*, Charlotte D. Vavourakis\*, James E. Barrett, Gerlinde Karbon, Andreas Villunger, Jiangrong Wang, Karin Sundström, Joakim Dillner, Martin Widschwendter


\* contributed equally

\tableofcontents

\pagebreak

```{r, warning = FALSE, message=FALSE, fig.width=15, fig.height=8}
#-------- Cell composition in training (Fig 1A)
# dat_cin <- readRDS("../0-data/pheno_P1.Rds")
#
# cin.colours2 <- c("#00A08799","#3C548899", "#7E614899", "#F39B7F99", "#DC000099")
#
# dat <- dat_cin %>%
#   pivot_longer(grep("hepidish", colnames(dat_cin)),
#                names_to = "celltype",
#                values_to = "value")
# 
# dat$set <- dat$split
# 
# celltypes <- dat %>%
#   filter(set == "training set") %>%
#   mutate(type = case_when(type == "Control_NEG" ~ "HPV-",
#                           type == "Control_POS" ~ "HPV+"),
#          celltype = case_when(celltype == "hepidish_NK" ~ "NK cells",
#                               celltype == "hepidish_Neutro" ~ "Neutrophils",
#                               celltype == "hepidish_Mono" ~ "Monocytes",
#                               celltype == "hepidish_Fib" ~ "Fibroblasts",
#                               celltype == "hepidish_Epi" ~ "Epithelial cells",
#                               celltype == "hepidish_Eosino" ~ "Eosinophils",
#                               celltype == "hepidish_CD8T" ~ "CD8+ T cells",
#                               celltype == "hepidish_CD4T" ~ "CD4+ T cells",
#                               celltype == "hepidish_B" ~ "B cells")) %>%
#   mutate(celltype = factor(celltype, levels = c("Eosinophils",
#                                                 "CD8+ T cells", 
#                                                 "CD4+ T cells",
#                                                 "B cells",
#                                                 "NK cells",
#                                                 "Monocytes",
#                                                 "Fibroblasts",
#                                                 "Neutrophils",
#                                                 "Epithelial cells"))) %>%
#   ggplot(aes(x = celltype,
#              y = value,
#              fill = type)) +
#   geom_boxplot() +
#   coord_flip() +
#   stat_compare_means(method = "wilcox.test",
#                      aes(label = paste0("p=", ..p.format..)),
#                      label.y = 0.9,
#                      colour = "gray40",
#                      size = 2.75,) +
#   theme_classic() +
#   theme(legend.position = "top",
#         legend.title = element_blank(),
#         axis.ticks = element_blank()) +
#   scale_fill_manual(values = cin.colours2[c(1,2)],
#                     aesthetics = "fill") +
#   xlab("") +
#   ylab("Proportion")
#
#-------- DELTA BETA histogram plot (Fig 1A)
# dat <- readRDS("../0-data/delta-beta.Rds")
# 
# cin.colours2 <- c("#00A08799","#3C548899", "#7E614899", "#F39B7F99", "#DC000099")
# 
# plot <- ggplot(dat,aes(x=db))+ 
#   geom_histogram(data=subset(dat,Tissue=='Epithelial'),
#                  aes(y=..count../sum(..count..),
#                      fill=Tissue),
#                  alpha=0.7,
#                  binwidth=0.001)+
#   geom_histogram(data=subset(dat,Tissue=='Immune'),
#                  aes(y=..count../sum(..count..),
#                      fill=Tissue),
#                  alpha=0.6,
#                  binwidth=0.001)+
#   scale_fill_manual(name="Cell type",
#                     values=cin.colours2[c(4,2)],
#                     labels=c("Epithelial","Immune")) +
#   xlim(-0.05,0.05) +
#   theme_minimal() +
#   xlab("Delta-beta estimate") +
#   ylab("Frequency") +
#   theme(panel.grid = element_blank(),
#         legend.position = "top")

#-------- Training - N (Figure 1A)
# readRDS("../0-data/n_training.Rds")
# plot <- n_tr %>%
#   ggplot(aes(x = n_seq,
#              y = oob_auc)) +
#   geom_line(size = 1,
#             colour = cin.colours2[1]) +
#   geom_point(aes(x = 500,
#                  y = 0.8384676),
#              size = 2,
#              shape = 4) +
#   theme(panel.grid = element_blank(),
#         panel.background = element_blank()) +
#   coord_cartesian(xlim = c(0,5000)) +
#   xlab("Number of top differentially methylated CpGs\nused as input") +
#   ylab("Out-of-bag AUC estimate")



# final output can also be used from analysis-pipeline/15-output

## load WID-HPV calculated for CIN data, alongside phenotypic information
dat_cin <- readRDS("../0-data/pheno_P1.Rds")

# adjust WID_HPV for age and ic
fit <- lm(WID_HPV ~ age + ic, data = dat_cin[dat_cin$type=="Control_NEG" & dat_cin$split == "hold out test set",])
dat_cin$WID_HPV_adj <- dat_cin$WID_HPV - predict(fit, newdata = dat_cin)

# arrange factor "type" (sample groups) for uniform plotting
dat_cin$type <- factor(dat_cin$type, levels = c("Control_NEG", "Control_POS", "CIN1", "CIN2", "CIN3+"))

type.labs <- c("HPV-/cyt-", "HPV+/cyt-", "CIN1", "CIN2", "CIN3+")
names(type.labs) <- c("Control_NEG", "Control_POS", "CIN1", "CIN2", "CIN3+")
#cin.colours <- c("#00A087FF", "#3C5488FF", "#7E6148FF", "#F39B7FFF", "#DC0000FF")
cin.colours2 <- c("#00A08799","#3C548899", "#7E614899", "#F39B7F99", "#DC000099")
#names(cin.colours) <- c("Control_NEG", "Control_POS", "CIN1", "CIN2", "CIN3+")
names(cin.colours2) <- c("Control_NEG", "Control_POS", "CIN1", "CIN2", "CIN3+")
type.labs.3 <- c("HPV+/cyt-", "CIN1", "CIN2", "CIN3+") #HPV-v/cyt-ve removed
names(type.labs.3) <- c("Control_POS", "CIN1", "CIN2", "CIN3+")
cin.colours3 <- c("#3C548899", "#7E614899", "#F39B7F99", "#DC000099")
names(cin.colours3) <- c("Control_POS", "CIN1", "CIN2", "CIN3+")

# # subplot experimental design
# p1 <- ggplot() + geom_blank() +theme_bw()
# 
# # subplot performance:
# # 1. retain only data from cyt-ve women used to validate the WID-HPV;
# # 2. add dichotom var for age;
# # 3. set case/control
# # 4. sort factor levels (dichot var)
# # 5. sort type levels
# 
# dat2ROC <- dat_cin %>% 
#   filter(split == "hold out test set" & type %in% c("Control_NEG","Control_POS")) %>%
#   mutate(dichotom_var = case_when(age < 30 ~ "below 30",age >= 30 ~"above 30"), 
#          type2 = case_when(type == "Control_POS"~ "case",type == "Control_NEG" ~"control")) %>% 
#   mutate(dichotom_var = factor(dichotom_var, levels = c("below 30","above 30")),
#          type2 = factor(type2, levels = c("control","case"))) %>%
#   droplevels() 
# 
# library(RColorBrewer)
# # Plot 2: ROC (by age)
# p2 <- plot_roc_dichot(dat2ROC$type2,
#                       dat2ROC$WID_HPV_adj,
#                       dat2ROC$dichotom_var,
#                       col2 = "coral3",
#                       col3 = "burlywood3")
# 
# ggsave(p2, file = "../../plots-ch/auc_hpv.pdf",
#        width = 3.5, height = 3.5)
# 
# # subplot violins ctrls -> CIN3+
# type.labs[names(type.labs) == "Control_NEG"] <- "HPV-"
# type.labs[names(type.labs) == "Control_POS"] <- "HPV+"
# 
# HPV+/HPV-
# tmp <- dat_cin[dat_cin$split != "training set",]
# # HPV+/HPV-
# ref <- tmp$type == "Control_NEG"
# comp <- tmp$type == "Control_POS"
# hpvpos <- signif(p.adjust(wilcox.test(tmp[ref,]$WID_HPV_adj,
#                           tmp[comp,]$WID_HPV_adj)$p.value,
#                        n = 5,
#                        method = "bonferroni"), 2)
# 
# comp <- tmp$type == "CIN1"
# cin1 <- signif(p.adjust(wilcox.test(tmp[ref,]$WID_HPV_adj,
#                           tmp[comp,]$WID_HPV_adj)$p.value,
#                  n = 5,
#                  method = "bonferroni"), 2)
# 
# 
# comp <- tmp$type == "CIN2"
# cin2 <- signif(p.adjust(wilcox.test(tmp[ref,]$WID_HPV_adj,
#                     tmp[comp,]$WID_HPV_adj)$p.value,
#                  n = 5,
#                  method = "bonferroni"), 2)
# 
# comp <- tmp$type == "CIN3+"
# cin3 <- p.adjust(wilcox.test(tmp[ref,]$WID_HPV_adj,
#                     tmp[comp,]$WID_HPV_adj)$p.value,
#                  n = 5,
#                  method = "bonferroni")
# 
# # All + v CIN3+
# ref <- tmp$type %in% c("Control_POS", "CIN1", "CIN2")
# comp <- tmp$type == "CIN3+"
# cin3_all <- signif(p.adjust(wilcox.test(tmp[ref,]$WID_HPV_adj,
#                     tmp[comp,]$WID_HPV_adj)$p.value,
#                     n = 5,
#                     method = "bonferroni"),2)
# 
# 
# 
# 
# p3 <- tmp %>%
#   filter(split != "training set") %>%
#   ggplot(aes(x = type,
#              y = WID_HPV_adj,
#              fill=type)) +
#   geom_half_boxplot(width=0.2)+
#   geom_half_violin() +
#   geom_half_point(transformation = position_jitter(),size=0.3,aes(color=type)) +
#   ylab("WID-HPV") +
#   xlab("") +
#   scale_fill_manual(name="",labels=type.labs,values=cin.colours2) +
#   scale_color_manual(name="",labels=type.labs,values=cin.colours2) +
#   geom_hline(yintercept=0,linetype="dashed",color = "black") +
#   theme_classic() +
#   theme(axis.text.x=element_blank(),
#         axis.ticks.x=element_blank(),
#         legend.position = "top") +
#   annotate("text",
#            x = 2,
#            y = 3.75,
#            label = paste0("p = ", hpvpos),
#            size = 3.2,
#            colour = "gray40") +
#   annotate("text",
#            x = 3,
#            y = 3.75,
#            label = paste0("p = ", cin1),
#            size = 3.2,
#            colour = "gray40") +
#   annotate("text",
#            x = 4,
#            y = 3.75,
#            label = paste0("p = ", cin2),
#            size = 3.2,
#            colour = "gray40") +
#   annotate("text",
#            x = 5,
#            y = 3.75,
#            label = paste0("p = ", cin3),
#            size = 3.2,
#            colour = "gray40")
# 
# p3 <- tmp %>%
#   filter(split != "training set") %>%
#   ggplot(aes(x = type,
#              y = WID_HPV_adj,
#              fill=type)) +
#   geom_half_boxplot(width=0.2)+
#   geom_half_violin() +
#   geom_half_point(transformation = position_jitter(),size=0.3,aes(color=type)) +
#   ylab("WID-HPV") +
#   xlab("") +
#   scale_fill_manual(name="",labels=type.labs,values=cin.colours2) +
#   scale_color_manual(name="",labels=type.labs,values=cin.colours2) +
#   geom_hline(yintercept=0,linetype="dashed",color = "black") +
#   theme_classic() +
#   theme(axis.text.x=element_blank(),
#         axis.ticks.x=element_blank(),
#         legend.position = "top") +
#   annotate("segment",
#            x = c(2, 4, 2, 3, 3, 5),
#            xend = c(2, 4, 4, 3, 5, 5),
#            y = c(3.5, 3.5, 3.7, 3.7, 4, 4),
#            yend = c(3.7, 3.7, 3.7, 4, 4, 3.2),
#            size = 0.5) +
#   annotate("text",
#            x = 4,
#            y = 4.2,
#            label = paste0("p = ", cin3_all),
#            size = 3.1,
#            colour = "gray40")
# 
# 
# # 
# ggsave(p3, file = "../0-data/figures/hpv-cin.pdf",
#        width = 4.5, height = 3.5) 
# 
# # plot with HPV subtypes (now in supplementary)? more ROCs?
# 
# p4 <- ggplot() + geom_blank() + theme_bw() 
# 
# # print full plot
# layout <- "
# AAA
# BCD
# "
# 
# plot <- p1 + p2 +p3 + p4 + plot_annotation(tag_levels = 'a') + plot_layout(design = layout)
# 
# plot

```

\includegraphics[trim=0 0cm 0 0, clip]{../0-data/figures/Fig_1.pdf}

\subsection{Figure 1. WID-HPV index development and performance in relation to HPV status and cervical intraepithelial neoplasia.}

\pagebreak

```{r, warning = FALSE, message=FALSE, fig.width=15,fig.height=15}

# data for this plot will be generate in analysis-pipeline/8-output

# load and arrange GSEA data for bubble plots
table <- readRDS("../0-data/GSEA_training_WID_HPV/DMPs_vs_all_result_GO.Rds")
table <- table %>% filter(FDR<0.05) %>% rownames_to_column(var="ID") %>% droplevels # keep only significant pathways

# number of DMPs in all genes
gene_count <- readRDS("../0-data/GSEA_training_WID_HPV/DMP_count_siggenes_GO.Rds")

#format dataframes for plotting
gene_count <- gene_count %>% rownames_to_column(var="ID")
gene_count$DMP_count <- as.numeric(gene_count$DMP_count)
table <- table %>% 
  dplyr::rename(Category=ONTOLOGY,Term=TERM,Genes=SigGenesInSet, pval=P.DE, adj_pval=FDR)
circ <- circle_dat(table, gene_count)

# add number of DMPs per term to circ
count2 <- c()#init
terms <- unique(circ$ID)
for (i in 1:length(terms)){
  co <- sum(circ$DMP_count[circ$ID==terms[i]])
  count2 <- c(count2,co)
}
df <- data.frame(ID=terms, count2=count2)
circ <- full_join(circ,df)#combine data

# make a bubble plot
circ$count <- as.numeric(circ$count)
circ$pval_log <- -log(circ$pval,10)
circ$adj_pval_log <- -log(circ$adj_pval, 10)
circ$term <- paste0(circ$term," (", circ$ID,")")

# subplot go- BB
p1 <- circ %>%
  filter(category=="BP") %>% droplevels()%>%
  arrange(desc(adj_pval_log)) %>%
  ggplot(aes(x=count, y=count2, size=adj_pval_log, fill=term)) +
    geom_point(alpha=0.5, shape=21, color="#7A0177") +
    scale_size(range = c(min(circ$adj_pval_log), 2*max(circ$adj_pval_log))) +
  scale_fill_brewer(palette = "RdPu", direction=-1) +
  guides(size=guide_legend(ncol=3))+
  theme_minimal()+
  labs(size="-log(adj. p-val)", fill="") +
  xlab("Number of genes in term")+
  ylab("Number of DMPs in term")+
  ggtitle('Enriched GO terms, Biological process')

# subplot go - CC
p2 <- circ %>%
  filter(category=="CC") %>% droplevels()%>%
  arrange(desc(adj_pval_log)) %>%
  ggplot(aes(x=count, y=count2, size=adj_pval_log, fill=term)) +
    geom_point(alpha=0.5, shape=21, color="#005A32") +
    scale_size(range = c(min(circ$adj_pval_log), 2*max(circ$adj_pval_log))) +
  scale_fill_brewer(palette = "Greens", direction=-1) +
  guides(size=guide_legend(ncol=3))+
  theme_minimal()+
  labs(size="-log(adj. p-val)", fill="") +
  xlab("Number of genes in term")+
  ylab("Number of DMPs in term")+
  ggtitle('Enriched GO terms, Cellular component')
# note that membrane-enclosed lumen (GO:0031974),organelle lumen (GO:0043233),intracellular organelle lumen (GO:0070013) completely overlap

# subplot chord
table_BP <- table %>% filter(Category=="BP") %>% droplevels()
genes_BP <- c() #init
for (i in 1:length(table_BP$Genes)){
  out <- unlist(strsplit(table_BP$Genes[i],","))
  genes_BP <- c(genes_BP,out)
}
genes_BP <- unique(genes_BP)
gene_count_BP <- gene_count %>% filter(ID %in% genes_BP)
circ_BP <- circle_dat(table_BP, gene_count_BP)

# make matrix for chord plot
N <- length(table_BP$Term) + 1
M <- gene_count_BP[,1] 
mat <- matrix(0, ncol = N, nrow = length(M))
rownames(mat) <- M
colnames(mat) <- c(table_BP$ID, 'DMPcount') 
mat[,N] <- gene_count_BP[,2]
for (p in 1:(N-1)){
  dat <- subset(circ_BP, ID == table_BP$ID[p])
  for (g in 1:length(M)) mat[g, p] <- ifelse(M[g]%in%dat$genes, 1, 0)
    }
chord <- mat

p3 <- GOChord(chord,space = 0.02, limit= c(7,0) , gene.order = 'DMPcount', gene.space = 0.25, gene.size=3) # displaying only genes in minimum 7 categories, note that this function still doesn't work for setting limits higher on DMPcount :/
p3 <- p3 + ggtitle("Genes encompassing minimum 7 enriched GO terms, Biological process")


# subplot heatmap
## read in pre-calculated delta-beta values
db <- readRDS("../0-data/GSEA_training_WID_HPV/deltabeta_train_dmp.Rds")
annot_DMP_mM <- readRDS("../0-data/GSEA_training_WID_HPV/DMPs_annot_missMethyl.Rds")
cpg_8386_annot <- readRDS("../0-data/GSEA_training_WID_HPV/cpg_8386_annot.Rds") #need this file for CpG context information


## filter out genes and CpGs from GO term of interest
genes_GO_interest <- unlist(strsplit(table$Genes[table$Term=="positive regulation of DNA repair"],","))# 43 genes in term
annot_CpGs_GO_interest <- annot_DMP_mM %>% filter(alias %in% genes_GO_interest) #62 CpGs in 43 genes

## start with arranging factor gene, cause this determines order CpGs in final plot
gene <- annot_CpGs_GO_interest %>% dplyr::select(cpg,alias)
gene <- gene[with(gene, order(alias)),]
gene$alias <- factor(gene$alias, levels=c(unique(gene$alias)))
order_of_plot <- c(gene$cpg)
gene <- gene$alias

## make a matrix of the delta-beta values training set (main heatmap)
mat1 <- db %>% dplyr::select(c(db_epithelial,db_immune,db_mean)) %>% filter(row.names(db) %in% unique(annot_CpGs_GO_interest$cpg))
mat1 <- mat1[match(order_of_plot, row.names(mat1)),] #match order CpGs
#identical(row.names(mat1), order_of_plot) #sanity-check

## factor cgi 
cgi <- cpg_8386_annot %>% filter(Name %in% row.names(mat1)) %>% dplyr::select(Name,Relation_to_Island) %>% droplevels()
cgi$Relation_to_Island <- factor(cgi$Relation_to_Island, levels=c("N_Shelf","N_Shore","Island","S_Shore","S_Shelf","OpenSea"))
cgi <- cgi[match(order_of_plot, cgi$Name),] #match order CpGs
#identical(order_of_plot, as.character(cgi$Name)) #sanity-check
cgi <- cgi$Relation_to_Island

### factor GO term - could also add the occurence of these genes in other GO terms

### colors
cgi_col = c("Island" = magma(20)[5] ,"OpenSea" = magma(20)[18], "S_Shelf" = magma(20)[13],  "S_Shore" = magma(20)[9], "N_Shelf" = magma(20)[14],  "N_Shore" = magma(20)[10])
#col_fun = colorRamp2(c(-0.05, 0, 0.15), c("darkblue", "white", "darkred"))

### heatmap
ht_list =
  Heatmap(mat1, col = viridis(100), name = "Delta-beta", row_names_side = "left", row_names_gp = gpar(fontsize = 8), cluster_rows = FALSE, show_column_dend = FALSE, column_names_rot = 60, column_names_centered = TRUE,row_split = c(gene), row_title_rot = 0, row_title_gp = gpar(fontsize=8), cluster_row_slices=FALSE, border=TRUE, column_title = "Genes in term \n Positive Regulation of DNA repair") + 
  Heatmap(cgi, name = "CGI",  width = unit(5, "mm"), cluster_rows=FALSE,
    col = cgi_col, column_names_rot = 0, column_names_centered = TRUE)

p4 <- grid2grob(draw(ht_list, merge_legends = TRUE, heatmap_legend_side = "right"))

# print full plot
layout <- "
AAADD
BBBDD
CCCDD
CCCDD
"

plot <- p1 + p2 +p3 + p4 + plot_annotation(tag_levels = 'a') + plot_layout(design = layout)

plot


```

\subsection{Figure 2. Differential methylation and gene ontology analysis in HPV-infected and -uninfected samples.}

\pagebreak

```{r, warning = FALSE, message=FALSE, fig.width=10,fig.height=8}

dat_cin_notrain <- dat_cin %>% filter(split != "training set") %>% droplevels()

# correlation WID-HPV and pcgtage
p1 <- dat_cin_notrain %>%
  filter(type %in% c("Control_POS","CIN1","CIN2","CIN3+")) %>%
  droplevels() %>%
  ggplot(aes(x = pcgtage,
             y = WID_HPV_adj)) +
  geom_point(aes(color=type), size=0.6) +
  scale_color_manual(name="",labels=type.labs.3,values=cin.colours3) +
  scale_x_continuous(trans="log10") +
  theme_classic() +
  theme(panel.grid = element_blank()) +
  stat_cor(cor.coef.name ="tau", method="kendall", label.x.npc = "middle", size=3, label.y=2.5) +
  stat_cor(cor.coef.name ="rho", method="spearman", label.x.npc = "middle", size=3,label.y=2) +
  labs(y = "WID-HPV", x = "log(pcgtAge)", colour="")

# apoptosis hallmark genes
ref <- dat_cin_notrain$type %in% c("Control_POS", "CIN1", "CIN2")
comp <- dat_cin_notrain$type == "CIN3+"

cin3_all <- signif(wilcox.test(dat_cin_notrain[ref,]$median_methylation_apoptosis,
                               dat_cin_notrain[comp,]$median_methylation_apoptosis)$p.value,2)

p2 <- dat_cin_notrain %>%
  ggplot(aes(x = type,
             y = median_methylation_apoptosis, fill=type)) +
  geom_half_boxplot(width=0.2)+
  geom_half_violin() +
  geom_half_point(transformation = position_jitter(), nudge=0.05, size=0.3,aes(color=type)) +
  ylab("Median methylation\n(apoptosis gene promoters)") +
  xlab("") +
  scale_x_discrete(labels = type.labs) +
  scale_fill_manual(name="", labels=type.labs, values=cin.colours2)  +
  theme_classic() +
  scale_color_manual(name="", labels=type.labs, values=cin.colours2) +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())  +
  annotate("segment",
           x = c(2, 4, 2, 3, 3, 5),
           xend = c(2, 4, 4, 3, 5, 5),
           y = c(0.036, 0.036, 0.038, 0.038, 0.045, 0.045),
           yend = c(0.038, 0.038, 0.038, 0.045, 0.045, 0.044),
           size = 0.5) +
  annotate("text",
           x = 4,
           y = 0.046,
           label = paste0("p = ", cin3_all),
           size = 3.1,
           colour = "gray40")


# apoptosis index CIN data
ref <- dat_cin_notrain$type %in% c("Control_POS", "CIN1", "CIN2")
comp <- dat_cin_notrain$type == "CIN3+"
cin3_all <- signif(wilcox.test(dat_cin_notrain[ref,]$index_apoptosis,
                    dat_cin_notrain[comp,]$index_apoptosis)$p.value,2)

p3 <- dat_cin_notrain %>%
  ggplot(aes(x = type,
             y = index_apoptosis, fill=type)) +
  geom_half_boxplot(width=0.2)+
  geom_half_violin() +
  geom_half_point(transformation = position_jitter(), nudge=0.05, size=0.3,aes(color=type)) +
  ylab("Apoptosis index") +
  xlab("") +
  scale_x_discrete(labels = type.labs) +
  scale_fill_manual(name="", labels=type.labs, values=cin.colours2) +
  scale_color_manual(name="", labels=type.labs, values=cin.colours2) +
  theme_classic() +
  theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank()) +
  annotate("segment",
           x = c(2, 4, 2, 3, 3, 5),
           xend = c(2, 4, 4, 3, 5, 5),
           y = c(-6.25, -6, -5.5, -5.5, -5.25, -5.25),
           yend = c(-5.5, -5.5, -5.5, -5.25, -5.25, -5.5),
           size = 0.5) +
  annotate("text",
           x = 4,
           y = -4.75,
           label = paste0("p = ", cin3_all),
           size = 3.1,
           colour = "gray40")


# correlation WID-HPV and apoptosis index
p4 <- dat_cin_notrain %>%
  filter(type %in% c("Control_POS","CIN1","CIN2","CIN3+")) %>%
  droplevels() %>%
  ggplot(aes(x = index_apoptosis,
             y = WID_HPV_adj)) +
  geom_point(aes(colour=type), size=0.6) +
   scale_color_manual(name="",labels=type.labs.3,values=cin.colours3) +
  stat_cor(cor.coef.name ="tau", method="kendall", label.x = -17, size=3, label.y=2.5) +
  stat_cor(cor.coef.name ="rho", method="spearman", label.x = -17, size=3,label.y=2) +
  theme_classic() +
  labs(y = "WID-HPV", x = "Apoptosis index", colour="")

# print full plot
layout <- "
AB
CD
"

(plot <- p1 + p2 + p3 + p4 + plot_annotation(tag_levels = 'a') + plot_layout(design=layout, guides="collect"))

ggsave(plot, file = "../0-data/figures/Fig_2.pdf",
       width = 10, height = 8)
# p values can be superscripted in Illustrator for visual purposes.
```

\subsection{Figure 3. WID-HPV correlation with epigenetic replicative age and apoptosis.}
